Highly Available Transactions: Virtues and Limitations

### 2013-03-27 writeup/evaluation.tex
more writing

\begin{enumerate}
\item We revisit traditional database concurrency control with a focus
  on coordination costs and on high availability.
\item We examine the properties required by a realistic OLTP
  application patterned on TPC-C.
\item We perform a brief experimental evaluation of HAT versus non-HAT
  properties on public cloud infrastructure.
\end{enumerate}

@@

\vspace{.5em}\noindent\textbf{Serializablity} We begin by considering several
existing algorithms that provide serializability but are not highly
available. In summary, all algorithms in this
class~\cite{bernstein-book} require at least one round trip time (RTT)
before committing.

\vspace{.5em}\noindent\textit{Two-phase locking} (ignoring deadlocks) requires $T$ lock
operations and 1 unlock operation for a transaction of length $T$. If
all items are located on a single master server, this requires one RTT
(for requests originating from servers other than the master). If
items are located on $S$ ($S \leq T$) different servers, two-phase
locking requires, at minimum, $2(S-1)$ RTTs---one for lock and another
for unlock.

\vspace{.5em}\noindent\textit{Optimistic concurrency control} allows
reads from any server but transactions must be checked by a scheduler
before commit. With one scheduler per database, this requires one RTT
specifically for concurrency control. If schedulers are sharded by
database partition, a multi-partition transaction may require
contacting several different schedulers and, possibly, an atomic
commit protocol, which becomes substantially more
expensive.

\vspace{.5em}\noindent\textit{Deterministic transaction scheduling} is
effectively the inverse of optimistic concurrency control: a scheduler
first orders the transactions, then sends them to database replicas
for in-order execution. These techniques, pioneered in the context of
totally ordered atomic broadcast
systems~\cite{deterministic-scheduling}, have recently seen increased
attention~\cite{calvin} and effectively require communication overhead
similar to optimistic concurrency control. Finally,

\vspace{.5em}\noindent\textit{Multiversion concurrency control} allows readers to read older
versions of data items but, like OCC, must eventually be checked for
write-write conflicts, requiring a scheduler.

\vspace{.5em} Perhaps unsurprisingly, the reliance on a global total
order necessitates a minimum of one round-trip to a designated master
or coordination service for each of these classic algorithms. This is
simply a cost of providing serializability, and the cost of the
(minimum one) round trip will be determined by the deployment
environment, as we saw in Section~\ref{sec:motivation}.

\vspace{.5em}\noindent\textbf{Non-serializablity} While it is possible to use
several of the prior mechanisms for HAT systems---for example, a
multi-version storage system is a reasonable candidate for the basis
of a HAT database, albeit with write conflict checking
disabled---several techniques, like locking, are not suitable for
HATs. In fact, there are several mechanisms that do not provide
serializability but provide transactional semantics with varying
availability properties:

\vspace{.5em}\noindent\textit{Lock-based protocols} such as those used to provide weak
  isolation in Gray's original proposal~\cite{gray-isolation} require
  at least one RTT to the a lock server (or, alternatively, a majority
  quorum). In the presence of failure, distributed locking is
  challenging.

\vspace{.5em}\noindent The \textit{MDCC} commit protocol uses a Paxos
variant and escrow transactions to provide read committed (without
Lost Update) and Non-monotonic Snapshot Isolation with 1-2 RTTs but is
unavailable due to Lost Update~\cite{mdcc}.

\vspace{.5em}\noindent\textit{Parallel Snapshot Isolation} as
developed by the Walter system provides (unavailable) snapshot
isolation variant and, even when writes commute, requires 1 RTT to
enforce read recency~\cite{walter}. However, like other Snapshot
Isolation variants, PSI could be made highly available by giving up
recency requirements and Lost Update prevention.

\vspace{.5em}\noindent\textit\textbf{Eiger} provides read-only and
write-only transactions with item cut isolation and causal
consistency. It relies on sticky high availability with a linearizable
set of replicas within a datacenter. This provides low latency across
data centers. In the presence of server failure during write
transactions, reads and writes may become unavailable. Eiger's
predecessor \textit{COPS} provides similar semantics for read-only
transactions but does not suffer from unavailability as it does not
rely have a coordinator.

\vspace{.5em}\noindent\textit{Chan and Gray's Read-only transactions}
use a similar algorithm to Eiger to provide read-only transactions
with item cut isolation and causal consistency but client version
selection coordination is performed on reads instead of
writes~\cite{readonly}.

\vspace{.5em}\noindent\textit{Bolt-on Causal
  Consistency}~\cite{bolton} and \textit{Swift}~\cite{swift} can
provide read-only and write-only transactions with item cut isolation
and causal consistency with sticky high availability and both rely on
client-side caching.

\vspace{.5em}\noindent\textit{Bayou} provides several session
guarantees but, in its original implementation, may face unavailabilty
in the presence of failure~\cite{sessionguarantees}.

\vspace{.5em}\noindent\textit{Brantner's S3 Database} uses caching to
provide several session and transactional guarantees on top of cloud
infrastructure, often relying on linearizable queues and, accordingly,
are often unavailable~\cite{kraska-s3}

\vspace{.5em}

As we have seen, it is possible to implement many guarantees weaker
than serializability---including guarantees that are achievable with
high availability---and not achieve high availability. These
designs---though often unavailable---offer benefits unavailable to
HATs: for example, lock-based read committed, provide recency
guarantees in addition to read committed isolation! We expect that
further understanding the trade-offs between semantic strength,
performance, and storage and metadata requirements will remain an area
of active research for the foreseeable future.

>>>>

\begin{myenumerate}
\item We revisit traditional database concurrency control with a focus
  on coordination costs and on high availability.
\item We examine the properties required by a realistic OLTP
  application patterned on TPC-C.
\item We perform a brief experimental evaluation of HAT versus non-HAT
  properties on public cloud infrastructure.
\end{myenumerate}

@@

\vspace{.5em}\noindent\textbf{Serializablity} To establish a serial
order on transactions, all existing algorithms for achieving
serializability of general-purpose read-write transactions in a
distributed setting~\cite{bernstein-book} may require at least one
round trip time (RTT) before committing. As an example, traditional
two-phase locking for a transaction of length $T$ requires $T$
\texttt{lock} operations and at least one unlock operation.  Each of
these operations requires coordination, either with other database
replicas or with a lock service. If the coordination mechanism for
locking is unavailable, transactions cannot make progress. Similarly,
optimistic concurrency control requires contacting a validation
service, while deterministic transaction
scheduling~\cite{deterministic-scheduling} requires a
scheduler. Multi-version concurrency control fares somewhat better but
must still be checked for write conflicts.

Perhaps unsurprisingly, the reliance on a globally agreed total order
necessitates a minimum of one round-trip to a designated master or
coordination service for each of these classic algorithms.  This is
intuitively a consequence of serializability's lack of high
availability~\cite{davidson-survey} (or, as we have shown, reliance on
preventing Lost Update and Write Skew).  The cost of the (minimum one)
round trip will be determined by the deployment environment, as we saw
in Section~\ref{sec:motivation}.

\vspace{.5em}\noindent\textbf{Non-serializablity} Perhaps
interestingly, many existing distributed implementations of weak
isolation are not highly available. Lock-based proposals such as those
used to provide weak isolation in Gray's original
proposal~\cite{gray-isolation} do not degrade gracefully in the
presence of partial failure. However, lock-based protocols \textit{do}
provide recency guarantees. While multi-versioned storage systems
allow for a variety of transactional guarantees, we have not seen
proposals for truly weak isolation (e.g., non-``tentative update''
schemes) in this context. Chan and Gray's read-only transactions
provide read-only transactions with item-cut isolation, causal
consistency, and transactional atomicity but are unavailable in the
presence of coordinator failure~\cite{readonly}, similar to read-only
and write-only transactions more recently proposed by
Eiger~\cite{eiger}.  Bolt-on causal consistency~\cite{bolton},
Swift~\cite{swift}, Brantner's S3 database~\cite{kraska-s3}, and
Bayou~\cite{sessionguarantees} can all provide variants of the above
model with high availability, but none provide this HAT functionality
without modification. Similarly, the MDCC~\cite{mdcc} protocol offers
Read Committed isolation with Lost Update avoidance but is similarly
unavailable due to its reliance on preventing write conflicts.

As we have seen, it is possible to implement many guarantees weaker
than serializability---including guarantees that are achievable with
high availability---and not achieve high availability. We expect that
further understanding the trade-offs between semantic strength,
performance, and storage and metadata requirements will remain an area
of active research for the foreseeable future.

### 2013-03-27 writeup/hat-analysis.tex
more writing

Transactional atomicity is achievable in a HAT environment. To ensure
transactional atomicity, servers need to know when all of a
transaction's updates are present on their respective
replicas. However, servers do \textit{not} need to agree exactly when
all replicas are aware that all of a transaction's updates are present
on their respective replicas. This subtle distinction corresponds to
the difference between the distributed systems concepts of reliable
broadcast and consensus. The former requires that, if one server
reveals a write, then all servers reveal a write, while the latter
requires explicit agreement across servers (and is unavailable). We
omit a full discussion of the algorithm and instead refer the reader
to the extended version of this paper~\cite{hat-tr}.

@@

\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} stipulates
that a client's subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any total ordering on
transactions.

@@

We can achieve the above guarantees by forcing servers to wait to
reveal new writes until their respective dependencies are fulfilled on
all replicas. In the case of Monotonic Reads, servers should only
reveal new writes to a data item when all other servers responsible
for the item have also seen them. Otherwise, a client might read a
newer version, subsequently become partitioned from its prior server,
and contact a comparatively out-of-date server. In the case of
Monotonic Writes, each new write should become visible only when each
client's prior write is visible everywhere, along with Monotonic Reads
guarantees. For Writes Follow Reads, a servers should similarly wait
to reveal items until their dependencies are present everywhere.

Effectively, this mechanism requires that clients only read from a
globally agreed upon lower bound on the versions written to the
system. This is highly available as a client will never block
due to inability to find a server with a sufficiently up-to-date
version of a data item. However, it does not imply that transactions
will read their own writes or provide much confidence in making
forward progress through the version history in the presence of
partitions. The problem is that, if a server is partitioned, under the
highly available model, we cannot discount the possibility that an
unfortunate client will be forced to issue her next requests against
the partitioned server!

>>>>

Transactional atomicity is achievable in a HAT environment. To ensure
transactional atomicity, servers need to know when all of a
transaction's updates are present on their respective
replicas. However, replicas do not need to agree on when to reveal a
new value to clients. In distributed systems terminology,
transactional atomicity only requires reliable broadcast (with some
additional metadata for clients). We omit a full discussion of the
algorithm and instead refer the reader to the extended version of this
paper~\cite{hat-tr}.

@@

\vspace{.5em}\noindent\textit{\textbf{Monotonic reads}} requires that
a client's subsequent reads to a given object ``never return any
previous values''; reads from each item progress forward in a total
order. The ordering of reads should respect any total ordering on
transactions.

@@

We can achieve the above guarantees by forcing servers to wait to
reveal new writes until their respective dependencies are fulfilled on
all replicas. In the case of Monotonic Reads, servers should only
reveal new writes to a data item when all other servers responsible
for the item have also seen them. In the case of Monotonic Writes and
Writes Follow Reads, each new write should become visible only when
each write's dependencies are visible everywhere. This mechanism
effectively requires that all clients read from a globally agreed upon
lower bound on the versions written to the system. This is highly
available as a client will never block due to inability to find a
server with a sufficiently up-to-date version of a data item. However,
it does not imply that transactions will read their own writes or, in
the presence of partitions, make forward progress through the version
history. The problem is that, if a server becomes partitioned, under
the highly available model, we must handle the possibility that an
unfortunate client will be forced to issue her next requests against
the partitioned server!

### 2013-03-27 writeup/motivation.tex
more writing

Several recent studies quantify partition behavior more rigorously. A
2011 study of several Microsoft datacenters found a mean of 40.8
network link failures per day (95th percentile: 136), with a median
time to repair of around five minutes (and up to one week). Perhaps
surprisingly, provisioning redundant networks only reduces impact of
failures by up to 40\%, meaning network providers cannot easily
curtail partition behavior~\cite{sigcomm-dc}. A 2010 study of over 200
wide-area routers found an average of 16.2--302.0 failures per link
per year with an average annual downtime of 24--497 minutes per link
per year (95th percentile at least 34 hours)~\cite{sigcomm-wan}. In
HP's managed enterprise networks, WAN, LAN, and connectivity problems
account for 28.1\% of all customer support tickets while 39\% of
tickets relate to network hardware.  The median incident duration for
highest priority tickets ranges from 114--188 minutes and up to a full
day for all tickets~\cite{turner2012failure}. The incidence of network
failures is further confirmed by several prior studies showing for
example, on Sprint's WANs, a median time to repair between 2 and 1000
seconds, and a median time between failures of approximately 3000
seconds~\cite{ip-backbone-failures}, as well as frequent path routing
failures on the Internet~\cite{labovitz-failures}. Isolating,
quantifying, and designing for these network failures is an area of active
research in networking community~\cite{surviving-failures-bodik,
  uw-failure-networks}.

>>>>

Several recent studies quantify partition behavior more rigorously. A
2011 study of several Microsoft datacenters found a mean of 40.8
network link failures per day (95th percentile: 136), with a median
time to repair of around five minutes (and up to one week). Perhaps
surprisingly, provisioning redundant networks only reduces impact of
failures by up to 40\%, meaning network providers cannot easily
curtail partition behavior~\cite{sigcomm-dc}. A 2010 study of over 200
wide-area routers found an average of 16.2--302.0 failures per link
per year with an average annual downtime of 24--497 minutes per link
per year (95th percentile at least 34 hours)~\cite{sigcomm-wan}. In
HP's managed enterprise networks, WAN, LAN, and connectivity problems
account for 28.1\% of all customer support tickets while 39\% of
tickets relate to network hardware.  The median incident duration for
highest priority tickets ranges from 114--188 minutes and up to a full
day for all tickets~\cite{turner2012failure}. Other studies confirm
these results, showing median time between failures of approximately
3000 seconds with a median time to repair between 2 and 1000
seconds~\cite{ip-backbone-failures} as well as frequent path routing
failures on the Internet~\cite{labovitz-failures}. Isolating,
quantifying, and designing for these network failures is an area of
active research in networking
community~\cite{surviving-failures-bodik, uw-failure-networks}.
